---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---
This document will attempt to answer several questions as they relate to the [Activity Monitoring Date](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip);

1. What is the mean total number of steps taken per day?
2. What is the average daily activity?
3. Are there differences in activity patterns between weekdays and weekends?

## Loading and preprocessing the data
```{r echo=TRUE}
library(plyr)
data <- read.csv("activity/activity.csv")
data[, "date"] <- as.Date(data[, "date"], "%Y-%m-%d")
```


## What is mean total number of steps taken per day?
```{r}
meantotal <- ddply(data, 'date', summarize, total=sum(steps))
hist(meantotal$total, main="Total Number of Steps Taken Each Day", xlab="Steps")

meansteps <- mean(meantotal$total, na.rm=TRUE)
mediansteps <- median(meantotal$total, na.rm=TRUE)
```


The median total number of steps taken is `r mediansteps`, and the mean total number of steps taken is `r format(meansteps, digits=8)`.

## What is the average daily activity pattern?

```{r echo=TRUE}
meanavg <- ddply(data, 'interval', summarize, average=mean(steps, na.rm=TRUE))
plot(meanavg$interval, meanavg$average, type="l", xlab="Interval", ylab="Mean Steps", main="Mean Steps Time Series Plot Across Intervals" )
sorted <- arrange(meanavg, desc(average))
maxinterval <- sorted[1,"interval"]

```


The interval with the average maximum number of steps taken is `r maxinterval `.

## Imputing missing values

```{r echo=TRUE}

missing <- sum(is.na(data$steps))
total <- length(data$steps)

```

The number of intervals with missing data is `r missing` out of a total of `r total`. 

```{r echo=TRUE}

newdata <- data

for(i in 1:nrow(newdata)){
  datarow <- newdata[i,]
  if(is.na(datarow$steps)){
    newdata[i, "steps"] <- meanavg[meanavg$interval == datarow$interval, "average"]
  }
}

meantotalrev <- ddply(newdata, 'date', summarize, total=sum(steps))
hist(meantotalrev$total, main="Total Number of Steps Taken Each Day", xlab="Steps")

meanstepsrev <- mean(meantotalrev$total, na.rm=TRUE)
medianstepsrev <- median(meantotalrev$total, na.rm=TRUE)

```


The median total number of steps taken is `r format(medianstepsrev, digits=8)`, and the mean total number of steps taken is `r format(meanstepsrev, digits=8)`.

By replacing the missing values with the mean for that interval, we have shifted the median a small amount, while the mean has stayed the same.  As for the distribution we have increased the frequency of intervals that contain between 10,000 and 15,000 steps.

## Are there differences in activity patterns between weekdays and weekends?

```{r echo=TRUE}

newdata <- cbind(newdata, weekend = as.factor(ifelse(weekdays(newdata$date) %in% c("Saturday","Sunday"), "weekend", "weekday")))

meanavg <- ddply(newdata, c("interval", "weekend"), summarize, average=mean(steps, na.rm=TRUE))

library(lattice)
xyplot(average~interval|weekend, data=meanavg, layout=c(1,2), type="l", xlab="Interval", ylab="Number of steps")

```

All figures contained in this document have been saved in the PA1_template_files/figure-html/ folder, with the names unnamed-chunk-2-1.png (initial histogram), unnamed-chunk-3-1.png(mean step time series), unnamed-chunk-5-1.png (second histogram), and unnamed-chunk-6-1.png (panel plot weekend vs weekday).